Deep learning for quality control of surface physiographic fields using satellite Earth observations

### 2023-05-23 docs/submission/template.tex
extra plots, corrected

\begin{abstract}
About 2/3 of all densely populated areas (i.e. at least 300 inhabitants per km$^2$) around the globe are situated within a 9 km radius of a permanent waterbody (i.e. inland water or sea/ocean coast), since inland water sustains the vast majority of human activities. Water bodies exchange mass and energy with the atmosphere and need to be accurately simulated in numerical weather prediction and climate modelling as they strongly influence the lower boundary conditions such as skin temperatures, turbulent latent and sensible heat fluxes and moisture availability near the surface. All the non-ocean water (resolved and sub-grid lakes and coastal waters) are represented in the Integrated Forecasting System (IFS) of the European Centre for Medium-Range Weather Forecasts (ECMWF) model, by the Fresh-water Lake (FLake) parametrisation, which treats $\sim$ 1/3 of the land. It is a continuous enterprise to update the surface parametrization schemes and their input fields to better represent small-scale processes. It is, however, difficult to quickly determine both the accuracy of an updated parametrisation, and the added value gained for the purposes of numerical modelling. The aim of our work is to quickly and automatically assess the benefits of an updated lake parametrisation making use of a neural network regression model trained to simulate satellite observed surface skin temperatures. \hl{In order to rapidly assess accuracy of new surface physiography fields and if their use in the model increase predictability a neural network regression model (further in the text referred as VESPER) that can learn the mapping between a set of features x and targets y is constructed. In this case the features are the atmospheric and surface model fields (such as 2 metre temperature from ERA5 reanalysis) and the surface physiographic fields (such as orography and vegetation cover used to produce ERA5 reanalysis), see Table 1 for the full list of variables used. The target is the satellite land surface temperature (skin temperature from MODIS Aqua Day MYD11A1 v006 collection). VESPER can then make predictions about the skin temperature given a set of input variables (i.e. atmospheric and surface model fields, and surface physiographic fields). In turn, these predictions can then be compared against observations (i.e. satellite skin temperature).} We deploy this tool to determine the accuracy of recent upgrades to the FLake parametrisation, namely the improved permanent lake cover and the capacity to represent seasonally varying water bodies (i.e. ephemeral lakes). We show that for grid-cells where the lake fields have been updated, the prediction accuracy in the land surface temperature improves by 0.45 K on average, whilst for the subset of points where the lakes have been exchanged for bare ground (or vice versa) the improvement is 1.12 K. We also show that updates to the glacier cover improve further the prediction accuracy by 0.14 K. The inclusion of seasonal water is shown to be particularly effective for grid points which are highly time variable, generally improving the simulation accuracy by $\sim$1 K. The neural network regression model has proven to be useful and easily adaptable to assess unforeseen impacts of ancillary datasets, also detecting inappropriate changes of high vegetation to bare ground, which would lead to decreased the skin temperature simulation accuracy by 0.49 K, proving to be a valuable support to model development.
\end{abstract}

@@

\noindent It is a continuous enterprise to update the lake parametrization schemes and their input data fields to better represent small-scale surface processes. It is however challenging to accurately represent lakes in these parametrisations; the majority of lakes which are resolved at a 9km grid spacing have not had their morphology accurately measured, let alone monitored, whilst 28.9$\%$ of land and coastal cells are treated for sub-grid (i.e. covering half or less of a grid cell) water. When introducing an updated lake representation it is difficult apriori to determine the additional value gained through doing so. There are two key factors here:
\begin{itemize}
	\item Are the updated fields closer to reality?
	\item Do the updated fields increase the accuracy of the model predictions?
\end{itemize}
\textcolor{red}{The first point is straightforward; we want our parametrisation fields to better represent reality. If the lake depth of some lake is updated from 10m to 100m we want to be sure that 100m is closer to the true depth of the lake. For the second point, even if the updated fields are accurate, are they informative in the sense that they enable us to make more accurate predictions? For instance, the main target of lake parametrization is to reproduce lake surface water temperatures (and therefore evaporation rates). If a lake parametrisation scheme is updated to better represent different types of inland waterbodies, the time variability of inland waterbodies and/or the lake morphology fields use more in situ measurements, does this additional information allow for more accurate predictions of the lake surface water temperatures? Is it therefore worthwhile to update the parametrisation in this way? Since the resulting updated fields are ultimately used operationally, it is essential to ensure the accuracy of the fields and prevent any potential degradation or instability of the model. This problem of quickly and automatically verifying the accuracy and information gain of updated lake parametrisations is the aim of this work.} \newline

@@

\noindent This paper is organized as follows. In Section \ref{sec:2} we describe the construction of the VESPER tool - the raw input data, the processing steps and the construction of a neural network regressor. In Section \ref{sec:3} we then deploy VESPER to investigate and evaluate updated lake parametrisation fields. Discussion and concluding remarks are made in Sections \ref{sec:4} and \ref{sec:conclusion} respectively.

\section{Constructing VESPER}\label{sec:2}
In order to rapidly assess accuracy of new surface physiography fields and if their use in the model increase predictability a neural network regression model (further in the text referred as VESPER) that can learn the mapping between a set of features x and targets y is constructed. In this case the features are the atmospheric and surface model fields (such as 2 metre temperature from ERA5 reanalysis) and the surface physiographic fields (such as orography and vegetation cover used to produce ERA5 reanalysis), see Table 1 for the full list of variables used. The target is the satellite land surface temperature (skin temperature from MODIS Aqua Day MYD11A1 v006 collection).VESPER can then make predictions about the skin temperature given a set of input variables (i.e. atmospheric and surface model fields, and surface physiographic fields). In turn, these predictions can then be compared against observations (i.e. satellite skin temperature) and VESPER's accuracy evaluated. By varying the number/type and values of the input features x to VESPER and observing how the accuracy of its predictions change, some conclusions on if and how features can increase predictability of an actual atmospheric model. Moreover, by isolating geographic regions where the predictions get worse with new/updated surface physiographic fields, areas where these fields might be erroneous or not informative enough can be identified. Due to the inherent stochasticity of training a neural network regression model it is also possible for different models to settle in different local minimums i.e. the network variance. \newline

\noindent \hl{To understand the significance of this, every VESPER configuration was trained four times, each time with a different seed. The size of this variance is much smaller than the variance between different VESPER configurations} \textit{\textbf{TK:TBD}} \newline

@@

\noindent As focus of this study are lake related fields, and lakes occupy only 1.8\% of the Earth’s surface and are distributed very heterogeneously \citep{Choulga2014}, analysis of the results here was restricted to areas where there have been significant changes in the surface physiographic fields, i.e. a change in any of the surface field when going from V15 to V20 (and to V15X or V20X) of ≥ 10\% (≥ 0.1 for fractional fields); if lake or vegetation cover changed from 0.1 in V15 to 0.3 in V20 field set this change is classified as significant. Choice of ≥ 10 \% cut-off was adopted as it proofed to be a good trade off between having a sufficient number of grid points to inspect and the strength of the effect of changing the input field. As the cut-off \% increases less points are selected even though with more severe changes, whereas when the cut-off \% decreases more points are selected but it becomes more difficult to disentangle the change in the prediction accuracy from VESPER’s training noise. Alternative cut-off \% were briefly explored, but conclusions of the results remained broadly unchanged. All grid-cells selected for the analysis can be classified according to how the surface fields are updated when going from V15 to V20 (note that categories represent a systematic and consistent update across multiple related fields, and do not include any restrictions on other surface fields apart the ones mentioned): 
\begin{itemize}
	\item \textbf{Lake Updates}. The change in the lake cover \textit{cl} and lake depth \textit{dl} are significant, but the changes in ocean and glacier \textit{glm} fractions are not. This corresponds to grid-cells where lakes have been added or removed. Lake-Ground Updates is a sub-category where additional constraint that the change in the high/low vegetation fractions \textit{cvh/cvl} are not significant is in place. This then corresponds to the exchange of lakes for bare ground, or vice versa.
	\item \textbf{Vegetation Updates.} The change in the high vegetation fraction \textit{cvh} is significant, but the change in lake cover \textit{cl} is not significant. This corresponds to grid-cells where large features like forests and woodlands have been updated, exchanged for bare ground or low vegetation.
	\item \textbf{Glacier Updates}. The change in the glacier cover \textit{glm} is significant. This corresponds to any areas where the fraction of glacier ice has been updated.
\end{itemize}
The VESPER models is trained on 2016 data over the entire globe, and then used to predict LST for 2019. The training of a neural network is inherently stochastic - the same model trained twice with the same data can settle in different local optima and so make different predictions. To make our conclusions robust against this training noise, each VESPER model is in turn trained 4 times. For each MODIS ground truth we then have 4 LST predictions per model. To assess the changes of LST predictability due to the use of the updated surface physiographic fields instead of V15 field set (default) we compare the mean absolute error (MAE) between different VESPER models using the a simple metric $\delta_{\rm VM}$ is used:
\begin{equation} 
	\delta_{\rm VM} = \text{MAE}_{\rm VESPER\_VM} - \text{MAE}_{\rm VESPER\_V15} 
\end{equation}
where $M$ represents one of the field set versions V20, V20X or V15X, and $\text{MAE}$ is computed over the whole prediction period of 2019. In turn, the MAE is the error between the prediction of a VESPER model and the Aqua-MODIS LST. A negative $\delta_{\rm VM}$ indicates that the VESPER\_VM LST prediction is more accurate than the  VESPER\_V15 prediction, and vice versa.

@@

\subsection{Evaluation of updated lake fields}\label{sec:3}
To understand if there is a way to automatically and rapidly assess accuracy of updated and/or new surface physiography fields and if their use in the atmospheric model increase predictability a neural network regression model VESPER was trained over 2016 data and used to predict LST for 2019.

VESPER’s training noise was confirmed to be much smaller (\hl{estimated noise bias over all grid-cells for 2019 at ~31km resolution is 0.04K}) than differences in LST predictions by different VESPER configurations (\hl{degrees K}), so all changes in LST predictability can be attributed to the changes in surface physiographic fields.

As a first attempt lake related information is assessed, namely lake cover (and land-sea mask and glacier cover as they are used for lake cover generation) and lake mean depth, that were created from scratch using new up-to-date high-resolution input datasets (see Table \ref{tab:datasources}) for the V20 (and V20X) field set; other surface physiographic fields (see Table \ref{table:definitions}) were regenerated from the same input sources as in initial V15 field set but taking into account that lake related fields were changed. In cases when existing in V15 lake cover water was removed in V20, its place could fill high or low vegetation, glacier or bare ground. To analyse results grid-cells were divided into four categories, and each was investigated in detail (see Table \ref{tab:summary_table} for the results aggregated over the whole globe).

\begin{table*}
	\begin{tabular}{ccccccccc}
		\toprule
		\multirow{2}{*}{Category} & \multirow{2}{*}{Number of grid cells} & 	\multicolumn{4}{c}{Training noise of VESPER\_VM, $K$} &\multicolumn{3}{c}{$\delta_{\rm VM}$, $K$} \\  
		&&V15  & V15X & V20 & V20X & V15X &V20 & V20X  \\
		\hline 
		Lake&1631&0.09 & 0.09 & 0.08 & 0.09 &$-0.20$ &$-0.37$ & $-0.37$  \\
		Lake-Ground&546&0.16 & 0.17 & 0.15 & 0.09 &$-0.56$ &$-0.83$ & $-1.12$ \\
		Vegetation&58&0.25 & 0.31 & 0.36& 0.46 & $-4 \times 10^{-3}$&+0.04 & $+3 \times 10^{-4}$  \\
		Glacier&1057& 0.27& 0.30& 0.28 & 0.25 &$-9\times 10^{-3}$&$-0.22$ & $-0.28$  \\
		\bottomrule
	\end{tabular}
	\caption{Globally averaged differences VM between mean absolute error (MAE) of [VESPER\_VM LST predictions minus Aqua-MODIS LST] and MAE of [VESPER\_V15 LST predictions minus Aqua-MODIS LST] for 2019 at ~31km resolution (where M denotes V15X, V20, V20X field sets) per grid-cell category. Negative VM values indicate increase of LST predictability due to the use of the updated surface physiographic fields instead of V15 field set (default), positive VM values indicate decrease of LST predictability and suggest the presence of erroneous information in the surface physiographic fields. Training noise values are much smaller than the variance between different VESPER’s configurations, indicating that changes in LST predictability are mainly due to changes in the surface physiographic fields. \hl{TK: TBD}}
	\label{tab:categorisation}
\end{table*}

\subsubsection{Category: Lake updates}
\begin{figure}
	\includegraphics[width=\columnwidth]{categories_lakes_histogram}
	\caption{some caption} 
	\label{fig:example_figure_histogram}
\end{figure}

Lake Updates category shows significant improvements in LST predictability if using V20 field set instead of V15 (see Table \ref{tab:categorisation} and Figure \ref{fig:bitstring_100110}) – prediction accuracy increased globally (over 1631 grid-cells) on average by 0.45K \hl{(training noise in VESPERV15 XX, and in VESPERV20 XX)}, most notably in Australia and the Aral sea (two major regions where ephemeral lakes were removed and inland water distribution made up-to-date respectively, discussed in Surface physiographic fields section). In addition to the areas with a notable improvement in the prediction accuracy, there are some noteworthy regions where the predictions got worse (see red points in Figure \ref{fig:bitstring_100110}) suggesting inaccuracies or lack of information in the updated surface physiographic fields. A few of the most noteworthy grid-cells (see red points highlighted with green circles in Figure \ref{fig:bitstring_100110}) are:

@@

	\item \textbf{Tanzania}. There are two grid-cells at the centre and northern edge of Lake Natron, in Tanzania (lies to the south-east of Lake Victoria) where VESPERV20 LST predictions are less accurate than VESPERV15 \hl{(training noise in VESPERV15 XX and XX, and in VESPERV20 XX and XX, for the central and northern edge of the lake respectively)}. For the central point (see Figure \ref{fig:lakes1}) the lake fraction was increased from 0.04 in V15 to 0.39 in V20 field set; for the northern edge point lake fraction was also increased in V20 comparing to V15 field set along with a small decrease (∼0.1) in the low vegetation fraction. However, Lake Natron is a highly saline lake that often dries out, with high temperatures, high levels of evaporation and irregular rainfall. It is a highly complex and variable regime that is not well described by simply increasing fraction of permanent fresh water, and indeed results suggest that with current lake parametrization scheme it may be beneficial to keep the lake fraction low or introduce extra descriptor, e.g. salinity. 
	\item \textbf{Australia}. Grid-cell contains Lake Blanche, in South Australia (see Figure \ref{fig:lakes2}). Lake fraction was completely removed from 0.44 in V15 to 0.00 in V20 field set, along with lake depth reduction from 5.5m to 1.0m, and low vegetation fraction increase from 0.53 to 0.97. Whilst the removal of ephemeral water is generally accurate for Australia, for this grid-cell it lowers the predictability of VESPERV20 comparing to VESPERV15 \hl{(training noise in VESPERV15 XX, and in VESPERV20 XX)}. Lake Blanche is a salt lake that lies below sea level within a wetlands system (so will retain some surface water or soil moisture which will influence the temperature response), fairly devoid of any obvious vegetation. V20 field set describes grid-cell as above sea level (same in V15 orography fields) completely dry region covered with short grass (i.e. low vegetation), which is not an accurate description resulting in worse predictions.
	–\item \textbf{Salt Lake City, North America.} Grid-cell lies within the Great Salt Lake Desert, just to the west of the Great Salt Lake, Utah, US. Lake fraction was completely removed from GTR 0.50 in V15 to 0.00 in V20 field set, grid-cell is fully covered with bare ground in V20 field set \hl{(training noise in VESPERV15 XX, and in VESPERV20 XX)}. Whilst this area primarily is bare ground, satellite imagery also suggests the presence of a presumably highly saline lake (see Figure \ref{fig:lakes3}); in addition area has a large degree of orography and high elevation (∼1300m) which probably further complicates the surface temperature response. A more accurate description that accounts for the seasonality of the surface water and the salinity is necessary here.
	\item  \textbf{Afghanistan}. Grid-cell lies in the south-west of Afghanistan, close to the border with Iran. Lake fraction was completely removed from 0.11 in V15 to 0.00 in V20 field set \hl{(training noise in VESPERV15 XX, and in VESPERV20 XX)}. However, this area in fact has an extensive network of mountain tributaries which feed an ephemeral lake (e.g. see Figure \ref{fig:lakes4}). Most likely there should be some inland water within a year, especially during the rainy season, so complete water removal might be an overcorrection.

>>>>

\begin{abstract}
About 2/3 of all densely populated areas (i.e. at least 300 inhabitants per km$^2$) around the globe are situated within a 9 km radius of a permanent waterbody (i.e. inland water or sea/ocean coast), since inland water sustains the vast majority of human activities. Water bodies exchange mass and energy with the atmosphere and need to be accurately simulated in numerical weather prediction and climate modelling as they strongly influence the lower boundary conditions such as skin temperatures, turbulent latent and sensible heat fluxes and moisture availability near the surface. All the non-ocean water (resolved and sub-grid lakes and coastal waters) are represented in the Integrated Forecasting System (IFS) of the European Centre for Medium-Range Weather Forecasts (ECMWF) model, by the Fresh-water Lake (FLake) parametrisation, which treats $\sim$ 1/3 of the land. It is a continuous enterprise to update the surface parametrization schemes and their input fields to better represent small-scale processes. It is, however, difficult to quickly determine both the accuracy of an updated parametrisation, and the added value gained for the purposes of numerical modelling. The aim of our work is to quickly and automatically assess the benefits of an updated lake parametrisation making use of a neural network regression model trained to simulate satellite observed surface skin temperatures. In order to rapidly assess accuracy of new surface physiography fields and if their use in the model increase predictability a neural network regression model (further in the text referred as VESPER) that can learn the mapping between a set of features x and targets y is constructed. In this case the features are the atmospheric and surface model fields (such as 2 metre temperature from ERA5 reanalysis) and the surface physiographic fields (such as orography and vegetation cover used to produce ERA5 reanalysis), see Table 1 for the full list of variables used. The target is the satellite land surface temperature (skin temperature from MODIS Aqua Day MYD11A1 v006 collection). VESPER can then make predictions about the skin temperature given a set of input variables (i.e. atmospheric and surface model fields, and surface physiographic fields). In turn, these predictions can then be compared against observations (i.e. satellite skin temperature). We deploy this tool to determine the accuracy of recent upgrades to the FLake parametrisation, namely the improved permanent lake cover and the capacity to represent seasonally varying water bodies (i.e. ephemeral lakes). We show that for grid-cells where the lake fields have been updated, the prediction accuracy in the land surface temperature improves by 0.45 K on average, whilst for the subset of points where the lakes have been exchanged for bare ground (or vice versa) the improvement is 1.12 K. We also show that updates to the glacier cover improve further the prediction accuracy by 0.14 K. The inclusion of seasonal water is shown to be particularly effective for grid points which are highly time variable, generally improving the simulation accuracy by $\sim$1 K. The neural network regression model has proven to be useful and easily adaptable to assess unforeseen impacts of ancillary datasets, also detecting inappropriate changes of high vegetation to bare ground, which would lead to decreased the skin temperature simulation accuracy by 0.49 K, proving to be a valuable support to model development.
\end{abstract}

@@

\noindent It is a continuous enterprise to update the lake parametrization schemes and their input data fields to better represent small-scale surface processes. It is however challenging to accurately represent lakes in these parametrisations; the majority of lakes which are resolved at a 9km grid spacing have not had their morphology accurately measured, let alone monitored, whilst 28.9$\%$ of land and coastal cells are treated for sub-grid (i.e. covering half or less of a grid cell) water. When introducing an updated lake representation it is difficult apriori to determine the additional value gained through doing so. There are two key factors here:
\begin{itemize}
	\item Are the updated fields closer to reality?
	\item Do the updated fields increase the accuracy of the model predictions?
\end{itemize}
The first point is straightforward; we want our parametrisation fields to better represent reality. If the lake depth of some lake is updated from 10m to 100m we want to be sure that 100m is closer to the true depth of the lake. For the second point, even if the updated fields are accurate, are they informative in the sense that they enable us to make more accurate predictions? For instance, the main target of lake parametrization is to reproduce lake surface water temperatures (and therefore evaporation rates). If a lake parametrisation scheme is updated to better represent different types of inland waterbodies, the time variability of inland waterbodies and/or the lake morphology fields use more in situ measurements, does this additional information allow for more accurate predictions of the lake surface water temperatures? Is it therefore worthwhile to update the parametrisation in this way? Since the resulting updated fields are ultimately used operationally, it is essential to ensure the accuracy of the fields and prevent any potential degradation or instability of the model. This problem of quickly and automatically verifying the accuracy and information gain of updated lake parametrisations is the aim of this work. \newline

@@

\noindent This paper is organized as follows. In Section \ref{sec:2} we describe the construction of the VESPER tool - the raw input data, the processing steps and the construction of a neural network regressor. In Section \ref{sec:3} we then deploy VESPER to investigate and evaluate updated lake parametrisation fields. Discussion and concluding remarks are made in Sections \ref{sec:discussion} and \ref{sec:conclusion} respectively.

\section{Constructing VESPER}\label{sec:2}
In order to rapidly assess accuracy of new surface physiography fields and if their use in the model increase predictability a neural network regression model (further in the text referred as VESPER) that can learn the mapping between a set of features x and targets y is constructed. In this case the features are the atmospheric and surface model fields (such as 2 metre temperature from ERA5 reanalysis) and the surface physiographic fields (such as orography and vegetation cover used to produce ERA5 reanalysis), see Table 1 for the full list of variables used. The target is the satellite land surface temperature (skin temperature from MODIS Aqua Day MYD11A1 v006 collection).VESPER can then make predictions about the skin temperature given a set of input variables (i.e. atmospheric and surface model fields, and surface physiographic fields). In turn, these predictions can then be compared against observations (i.e. satellite skin temperature) and VESPER's accuracy evaluated. By varying the number/type and values of the input features x to VESPER and observing how the accuracy of its predictions change, some conclusions on if and how features can increase predictability of an actual atmospheric model. Moreover, by isolating geographic regions where the predictions get worse with new/updated surface physiographic fields, areas where these fields might be erroneous or not informative enough can be identified. Due to the inherent stochasticity of training a neural network regression model it is also possible for different models to settle in different local minimums i.e. the network variance. To understand the significance of this, every VESPER configuration was trained four times, each time with a different random seed. \newline

@@

\noindent As focus of this study are lake related fields, and lakes occupy only 1.8\% of the Earth’s surface and are distributed very heterogeneously \citep{Choulga2014}, analysis of the results here was restricted to areas where there have been significant changes in the surface physiographic fields, i.e. a change in any of the surface field when going from V15 to V20 (and to V15X or V20X) of ≥ 10\% (≥ 0.1 for fractional fields); if lake or vegetation cover changed from 0.1 in V15 to 0.3 in V20 field set this change is classified as significant. Choice of ≥ 10 \% cut-off was adopted as it proofed to be a good trade off between having a sufficient number of grid points to inspect and the strength of the effect of changing the input field. As the cut-off \% increases less points are selected even though with more severe changes, whereas when the cut-off \% decreases more points are selected but it becomes more difficult to disentangle the change in the prediction accuracy from VESPER’s training noise. Alternative cut-off \% were briefly explored, but conclusions of the results remained broadly unchanged. All grid-cells selected for the analysis can be classified according to how the surface fields are updated when going from V15 to V20 (note that categories represent a systematic and consistent update across multiple related fields, and do not include any restrictions on other surface fields apart the ones mentioned): 
\begin{itemize}
	\item \textbf{Lake Updates}. The change in the lake cover \textit{cl} and lake depth \textit{dl} are significant, but the changes in ocean and glacier \textit{glm} fractions are not. This corresponds to grid-cells where lakes have been added or removed. Lake-Ground Updates is a sub-category where additional constraint that the change in the high/low vegetation fractions \textit{cvh/cvl} are not significant is in place. This then corresponds to the exchange of lakes for bare ground, or vice versa.
	\item \textbf{Vegetation Updates.} The change in the high vegetation fraction \textit{cvh} is significant, but the change in lake cover \textit{cl} is not significant. This corresponds to grid-cells where large features like forests and woodlands have been updated, exchanged for bare ground or low vegetation.
	\item \textbf{Glacier Updates}. The change in the glacier cover \textit{glm} is significant. This corresponds to any areas where the fraction of glacier ice has been updated.
\end{itemize}
The VESPER models is trained on 2016 data over the entire globe, and then used to predict LST for 2019. The training of a neural network is inherently stochastic - the same model trained twice with the same data can settle in different local optima and so make different predictions. To make our conclusions robust against this training noise, each VESPER model is in turn trained 4 times. For each MODIS ground truth we then have 4 LST predictions per model. We will frequency refer to a ``training noise" as the standard deviation $\sigma$ in the VESPER predictions for the same input fields i.e. each VESPER model will have a corresponding training noise $\sigma_{\rm VM}$. To assess the changes of LST predictability due to the use of the updated surface physiographic fields instead of V15 field set (default) we compare the mean absolute error (MAE) between different VESPER models using the a simple metric $\delta_{\rm VM}$ is used:
\begin{equation} 
	\delta_{\rm VM} = \text{MAE}_{\rm VESPER\_VM} - \text{MAE}_{\rm VESPER\_V15} 
\end{equation}
where $M$ represents one of the field set versions V20, V20X or V15X, and $\text{MAE}$ is computed over the whole prediction period of 2019. In turn, the MAE is the error between the prediction of a VESPER model and the Aqua-MODIS LST. A negative $\delta_{\rm VM}$ indicates that the VESPER\_VM LST prediction is more accurate than the  VESPER\_V15 prediction, and vice versa.

@@

\subsection{Evaluation of updated lake fields}\label{sec:3}
To understand if there is a way to automatically and rapidly assess accuracy of updated and/or new surface physiography fields and if their use in the atmospheric model increase predictability a neural network regression model VESPER was trained over 2016 data and used to predict LST for 2019. Generally VESPER’s training noise was confirmed to be  smaller than differences in LST predictions by different VESPER configurations so changes in LST predictability can be meaningfully attributed to the changes in surface physiographic fields. Particular situations where the training noise becomes significant are discussed below. \newline

As a first attempt lake related information is assessed, namely lake cover (and land-sea mask and glacier cover as they are used for lake cover generation) and lake mean depth, that were created from scratch using new up-to-date high-resolution input datasets (see Table \ref{tab:datasources}) for the V20 (and V20X) field set; other surface physiographic fields (see Table \ref{table:definitions}) were regenerated from the same input sources as in initial V15 field set but taking into account that lake related fields were changed. In cases when existing in V15 lake cover water was removed in V20, its place could fill high or low vegetation, glacier or bare ground. To analyse results grid-cells were divided into four categories, and each was investigated in detail (see Table \ref{tab:categorisation} for the results aggregated over the whole globe).

\begin{table*}
	\begin{tabular}{ccccccccc}
		\toprule
		\multirow{2}{*}{Category} & \multirow{2}{*}{Number of grid cells} & 	\multicolumn{4}{c}{Training noise of VESPER\_VM, $K$} &\multicolumn{3}{c}{$\delta_{\rm VM}$, $K$} \\  
		&&V15  & V15X & V20 & V20X & V15X &V20 & V20X  \\
		\hline 
		Lake&1631 & 0.07& 0.02& 0.02& 0.02&  -0.2 & -0.37&-0.37 \\
		Lake-Ground&546 & 0.15 &0.05& 0.04& 0.06 & -0.56 &-0.83& -0.84 \\
		Vegetation&58 & 0.04 &0.1& 0.15& 0.21&  -$4 \times 10^{-3}$&0.04 & -3 $\times 10^-4$ \\
		Glacier&1057 & 0.03& 0.08& 0.02& 0.06 & -0.01& -0.22& -0.28  \\
		\bottomrule
	\end{tabular}
	\caption{Globally averaged differences VM between mean absolute error (MAE) of [VESPER\_VM LST predictions minus Aqua-MODIS LST] and MAE of [VESPER\_V15 LST predictions minus Aqua-MODIS LST] for 2019 at ~31km resolution (where M denotes V15X, V20, V20X field sets) per grid-cell category. Negative VM values indicate increase of LST predictability due to the use of the updated surface physiographic fields instead of V15 field set (default), positive VM values indicate decrease of LST predictability and suggest the presence of erroneous information in the surface physiographic fields. Training noise values are much smaller than the variance between different VESPER’s configurations, indicating that changes in LST predictability are mainly due to changes in the surface physiographic fields.}
	\label{tab:categorisation}
\end{table*}	
\begin{figure}
	\includegraphics[width=\columnwidth]{global_shift_plot}
	\caption{Distribution of prediction errors in the LST, for each of the 4 grid point categories, for each iteration of V15 and V20. For Lake, Lake-ground and Glacier categories the improvement in V20 relative to V15 is much greater than the intrinsic model noise, with all V20 predictions outperforming all V15 predictions. For the Vegetation category the predictions of V15 and V20 are much more noisy and it is difficult to draw any conclusions for the category as a whole.} 
	\label{fig:global_shift_plot}
\end{figure}

\subsubsection{Category: Lake updates}
\begin{figure}
	\includegraphics[width=\columnwidth]{lakes_histogram_plot}
	\caption{Distribution of prediction errors in the LST for all grid points in the lakes category for V15 and V20.  Each prediction errors is in turn the average of 4 trained iterations of the VESPER model.} 
	\label{fig:example_figure_histogram}
\end{figure}

Lake Updates category shows significant improvements in LST predictability if using V20 field set instead of V15 – prediction accuracy increased globally (over 1631 grid-cells) on average by 0.37K. For the lakes category, the training noise in V20 was generally small $\sigma_{\rm V20} \sim 0.02$ K, with the V15 predictions a little more noisy with $\sigma_{\rm V15} \sim 0.07$ K, but this noise is much less than the improvement - as can be seen in Fig. \ref{fig:global_shift_plot} every V20 iteration significantly outperforms every V15 iteration. In Fig. \ref{fig:example_figure_histogram} we plot the distribution of the mean LST error (averaged across each of the 4 trained VESPER iterations) for all lake grid points, for both V15 and V20. Evidently the V20 field significantly improve the high tail behaviour relative to V15, as well as shifting the median of the distribution to lower errors. Particular regions where the V20 physiographic fields notably improved performance were in Australia and the Aral sea (Fig. \ref{fig:bitstring_100110}). These are two major regions where ephemeral lakes were removed and inland water distribution made up-to-date respectively, as discussed in Surface physiographic fields section. In addition to the areas with a notable improvement in the prediction accuracy, there are some noteworthy regions where the predictions got worse (see red points in Figure \ref{fig:bitstring_100110}) suggesting inaccuracies or lack of information in the updated surface physiographic fields. A few of the most noteworthy grid-cells (see red points highlighted with green circles in Figure \ref{fig:bitstring_100110}) are:

@@

	\item \textbf{Tanzania}. There are two grid-cells at the centre and northern edge of Lake Natron, in Tanzania (lies to the south-east of Lake Victoria) where VESPERV20 LST predictions are less accurate than VESPERV15. For the central point (see Figure \ref{fig:lakes1}) the lake fraction was increased from 0.04 in V15 to 0.39 in V20 field set; for the northern edge point lake fraction was also increased in V20 comparing to V15 field set along with a small decrease (∼0.1) in the low vegetation fraction. However, Lake Natron is a highly saline lake that often dries out, with high temperatures, high levels of evaporation and irregular rainfall. It is a highly complex and variable regime that is not well described by simply increasing fraction of permanent fresh water, and indeed results suggest that with current lake parametrization scheme it may be beneficial to keep the lake fraction low or introduce extra descriptor, e.g. salinity. 
	\item \textbf{Australia}. Grid-cell contains Lake Blanche, in South Australia (see Figure \ref{fig:lakes2}). Lake fraction was completely removed from 0.44 in V15 to 0.00 in V20 field set, along with lake depth reduction from 5.5m to 1.0m, and low vegetation fraction increase from 0.53 to 0.97. Whilst the removal of ephemeral water is generally accurate for Australia, for this grid-cell it lowers the predictability of VESPERV20 comparing to VESPERV15. However the shift is only minor $\delta_{V20} = +0.03$ and far outweighed by the training noise (0.83, 1.54 for V15 and V20 respectively) so it is difficult to draw any meaningful conclusions. Lake Blanche is a salt lake that lies below sea level within a wetlands system (so will retain some surface water or soil moisture which will influence the temperature response), fairly devoid of any obvious vegetation. V20 field set describes grid-cell as above sea level (same in V15 orography fields) completely dry region covered with short grass (i.e. low vegetation), which is not an accurate description which could contribute to the lack of an improvement in the predictions.
	–\item \textbf{Salt Lake City, North America.} Grid-cell lies within the Great Salt Lake Desert, just to the west of the Great Salt Lake, Utah, US. Lake fraction was completely removed from GTR 0.50 in V15 to 0.00 in V20 field set, grid-cell is fully covered with bare ground in V20 field set. Whilst the training noise is significant at this grid point (0.26 for V15, 0.92 for V20) is it smaller than the error metric ($\delta_{V20} = 2.92K$) and we can see from Fig \ref{fig:lake_sub_categories} that the V20 predictions consistently underperform the V15 predictions. Whilst this area primarily is bare ground, satellite imagery also suggests the presence of a presumably highly saline lake (see Figure \ref{fig:lakes3}); in addition area has a large degree of orography and high elevation (∼1300m) which probably further complicates the surface temperature response. A more accurate description that accounts for the seasonality of the surface water and the salinity is necessary here.
	\item  \textbf{Afghanistan}. Grid-cell lies in the south-west of Afghanistan, close to the border with Iran. Lake fraction was completely removed from 0.11 in V15 to 0.00 in V20 field set. However, this area in fact has an extensive network of mountain tributaries which feed an ephemeral lake (e.g. see Figure \ref{fig:lakes4}). Most likely there should be some inland water within a year, especially during the rainy season, so complete water removal might be an overcorrection.
	\item \textbf{Northern India.} Grid-cell lies in the state of Gujarat, India, close to the border with Pakistan. Lake fraction was increased from 0.59 in V15 to 0.71 in V20 field set, along with the lake depth increase from 2.58m to 3.76m. However, this point lies on a river delta within the Great Raan of Kutch, a large area of salt marshes (see Figure \ref{fig:lakes5}), known for having highly seasonal rainfall, with frequent flooding during the monsoon season and a long dry season. The surface itself also undulates with areas of higher sandy ground known as medaks, with greater levels of vegetation. It is evidently a complex and highly time variable area and additional static fraction of fresh water provided via V20 field set is not enough.
	\item  \textbf{Egypt}. Grid-cell contains a section of the Toshka Lakes, lies to the west of the River Nile in the south of Egypt. Lake fraction was completely removed from 0.36 in V15 to 0.00 in V20 field set, along with the lake depth decrease from 25m to 6m \hl{(training noise in VESPERV15 XX, and in VESPERV20 XX)}. However, whilst this is a very dry region, it contains part of Toshka Lakes, a collection of endorheic waterbodies newly formed (and growing) due to overflow from Lake Nasser (see Figure \ref{fig:lakes6}). These lakes are known to be highly time variable, with a periodic seasonality on top of the general increasing lake sizes, and the formation of surrounding wetlands. These lakes rapidly fill and dry out; during the training and validation years of VESPER (along with the decade before) lakes were mostly dry, whereas during the testing year they were filled. 
\end{itemize}
\noindent It is worth mentioning a vast area north-west of Canada where VESPERV20 slightly underperforms VESPERV15 \hl{(V20 averaged over the area is +0.02K; training noise in VESPERV15 XX, and in VESPERV20 XX)}, but it is hard to draw any definitive conclusions – whilst LST predictions for some grid-cells become more accurate, for some – less accurate. For these high latitude regions there is a large uncertainty for lake location (thermokarst lakes emerge due to permafrost thawing, have small area and can quickly disappear) and lake depth which influence water freezing/thawing dates and ice duration. In addition, observations in these regions are more uncertain and have greater errors due to increased cloud cover. Another interesting region is the north-eastern edge of the Caspian Sea, where VESPERV20 underperforms VESPERV15 (V20 averaged over four grid-cells is +0.65K; \hl{training noise in VESPERV15 XX, and in VESPERV20 XX)}. Location is the Astrakhan Nature Reserve (an extensive wetland). In general, lake fraction was reduced and vegetation fraction increased accordingly in V20 comparing to V15 field set. However, it seems that permanent fresh water fraction is not enough to accurately describe the region and additional monthly varying water cover and/or wetland cover should be introduced. \newline
